\chapter*{ABSTRACT}
\addcontentsline{toc}{chapter}{ABSTRACT}

Diploma thesis: 70 pages, 14 figures, 2 tables, 30 references, 2 appendices.

\textbf{Keywords:}
quantum dot, tight-binding method, superellipse, low-energy spectrum, surrogate approximation, Ridge regression, MLP, structured validation.

\textbf{Object of research} --- model superelliptic quantum dots represented as finite nanostructures on a square tight-binding lattice.

\textbf{Subject of research} --- the dependence of the low-energy spectral characteristics \(E_0\) and \(dE_1\) on size and boundary shape, and the physically motivated surrogate approximation of these quantities.

\textbf{Aim of the work} --- to study whether the low-energy spectrum of model superelliptic tight-binding quantum dots can be described by physically motivated confinement descriptors, and whether a small MLP provides a robust advantage over interpretable Ridge regression under structured validation.

\textbf{Methods.}
The work uses a square-lattice tight-binding Hamiltonian with direct spectral calculation in Kwant. Physical consistency was checked by a rectangular control calculation, the energy convention \(E_{\mathrm{kin}} = E_0 + 4\), \(1/a^2\) scaling, and a circular Bessel-scale check. Ridge regression with physically motivated descriptors was used as the main surrogate approximation, while a small MLP was used as a controlled nonlinear comparison. Model quality was evaluated with LOAO and LOARO structured validation, and Ridge residuals were examined using correlation diagnostics.

\textbf{Main results and significance.}
A calculated sample of 140 geometries was formed with \(n = \{1.2, 2.0, 3.0, 4.0\}\), \(a = \{24, 27, 30, 33, 36\}\), and \(r_{\mathrm{AR}} = \{0.67, 0.72, 0.78, 0.83, 0.89, 0.94, 1.0\}\). In the circular control case, the Bessel-scale relative error was about \(2.03\%\). The fixed-shape scaling check gave a maximum relative deviation of about \(2.12\%\). The MLP with physically motivated descriptors satisfied the success rule in 10 of 16 comparison cells: 3 of 8 under LOAO and 7 of 8 under LOARO. The pre-registered criterion for robust superiority was not met, so Ridge regression remains the preferred main model. At the same time, the physically motivated MLP outperformed the MLP using raw geometric parameters in 13 of 16 cells.

Absolute MAE values were measured in units of the hopping parameter \(t\). For \(E_0\), relative error was evaluated with respect to \(E_{\mathrm{kin}}\), while for \(dE_1\) it was evaluated with respect to the characteristic gap. For Ridge regression, the relative MAE was approximately 2.01--2.60\% of \(E_{\mathrm{kin}}\) for \(E_0\), and 0.78--1.15\% of \(dE_1\) for the gap. The hypothesis that Ridge residuals are globally explained by simple boundary-discretization diagnostics was not supported. A diagnostic signal was found only for \(n=1.2\): the correlation with sublattice imbalance has \(\rho \approx -0.416\), \(p \approx 0.013\), and is interpreted cautiously.

\textbf{Scope and limitations.}
The work demonstrates a controlled methodology for combining direct tight-binding spectral calculations of finite nanostructures with physically motivated surrogate approximation. The results apply only to the studied parameter grid, and \(n\) is treated as a fixed discrete class. No universal theory of shape, completed inverse design, or replacement of direct Kwant calculation is claimed. Material-specific use would require additional parameter calibration or DFT/OpenMX reference data.

\clearpage
